
# Introduction to R programming {#introR}

This notebook collects the scripts used for teaching in BIOF1001 for 
Introduction to R (1 hour teaching). You can get this Rmd file 
[here](https://github.com/StatBiomed/BMDatSci-book/blob/main/notebooks/chapter1-R/01-introR.Rmd)
(right click and "save link as" to download).

## Data types
During computing, all data are loaded / stored in memory (RAM) through a group 
of binary bits, usually 8 bits as a byte. In R language setting (similar to 
some other programming languages), there are a few commonly used data types 
that have specific number of bytes to use.

In R, you can use the `class()` and `typeof()` functions to check the class and 
data type of any variable. In total, R has five data types:

1. Numeric
2. Integers
3. Complex
4. Logical
5. Characters


### nemeric (or double)
The `numberic` is for numeric values, as the most common and the **default** 
data type. The `numberic` datatype saves values in double precision (double 
number of bytes in memory), so the type is also `double`.

```{r}
x <- c(1.0, 2.0, 5.0, 7.0)
x
class(x)
typeof(x)
```

### integer
The integer is another data type used for the set of all integers. You can use 
the capital ‘L’ notation as a suffix to specify a particular value as the 
integer data type. Also, you can convert a value into an integer type using the `as.integer()` function.

```{r}
y <- c(1L, 2L, 5L, 7L)
y
class(y)
typeof(y)
```

```{r}
# Assign a integer value to y
y = 5
 
# is y an integer?
print(is.integer(y))
```


### logical
In R, the `logical` data type takes either a value of true or false. A `logical` 
value is often generated when comparing between variables.

```{r}
z <- c(TRUE, TRUE, TRUE, FALSE)
z
typeof(z)
```


### character
In R, the `character` is a data type where you have all the alphabets and 
special characters. It stores `character` values or `strings`. Strings in R can 
contain alphabets, numbers, and symbols. The character type is usually denoted 
by wrapping the value inside single or double inverted commas.

```{r}
w <- c("aa", "bb", "5", "7")
w
typeof(w)
```

### Memeory usage
As we will see below, integer and logical uses only half of the memory of double (numeric) and character.
```{r}
x <- c(1.0, 2.0, 5.0, 7.0)
y <- c(1L, 2L, 5L, 7L)
z <- c(TRUE, TRUE, TRUE, FALSE)
w <- c("aa", "bb", "5", "7")

object.size(x)
```
```{r}
object.size(rep(x, 1000))
object.size(rep(y, 1000))
object.size(rep(z, 1000))
object.size(rep(w, 1000))
```


## Data structures
Data structure is one of most important features in programming. It involves 
how the data is organised and can be accessed and modified. Using an appropriate
data structure may largely improve the computing efficiency.

### Vector
Vector is a basic data structure in R. It contains elements in the **same** data
type (no matter double, integer, character or others). 

You can check the data type by using `typeof()` function and the length of the 
vector by `length()` function.

Since, a vector must have elements of the same type, this function will try and coerce elements to the same type, if they are different. Coercion is from lower 
to higher types from logical to integer to double to character.

More introduction see [here](https://www.datamentor.io/r-programming/vector/).

```{r}
x <- c(1, 2, 5, 7)
x
typeof(x)

x <- rep(3, 5)
x
typeof(x)

x <- 1:12 # integer
x
typeof(x)
```

### Matrix
Matrix is a two-dimensional data structure. It is in principle build based on 
vector but have more convenient built-in functions for computation.
It has row and column, both of which can also have name. To check the dimensions,
you can use the `dim()` function.

More introduction see [here](https://www.datamentor.io/r-programming/matrix/).

```{r}
A <- matrix(1:12, nrow=3)
A

B <- matrix(1:12, nrow=3, byrow=TRUE)
B

colnames(A) <- c("C1","C2","C3","C4")
rownames(A) <- c("R1","R2","R3")
A
```

#### Index Vector and Matrix
Two index vector, you can use `logical` or `integer`, or the element name if it
has.

In R, the index *starts from 1*, unlike most programming languages where index 
start from 0.

We can also use negative integers to return all elements except that those 
specified. But we cannot mix positive and negative integers while indexing and 
real numbers, if used, are truncated to integers.

**N.B.**, when using logical as index, be careful if the index length is 
different from the vector.


```{r}
x <- 1:12

x[3]
x[2:5]
x[c(2, 5, 6)]                   # index with integer

x[c(TRUE, FALSE, FALSE, TRUE)]  # index with logical value
```

Now, index matrix. Very similar as vector, but it has both row and column.
```{r}
A <- matrix(1:12, nrow=3)
colnames(A) <- c("C1","C2","C3","C4")
rownames(A) <- c("R1","R2","R3")

A[1, 2]
A[1, "C2"]

A[1, c(2, 3)]
A[1:2, c(2, 3)]
```

Single row or column matrix will become a vector, unless **using `drop=FALSE`**
```{r}
A[1, 2:4]
dim(A[1, 2:4])

A[1, 2:4, drop=FALSE]
dim(A[1, 2:4, drop=FALSE])
```

#### Modify values
```{r}
A[1, 2:4] <- c(-3, -5, 20)
A
```

### List
Different from `vector` that have all elements in the same data type, the `list`
data structure can have components of mixed data types.

More broadly, a `list` can contain a list of any data structure: value, vector, 
matrix, etc
```{r}
x <- list(2.5, TRUE, 1:3)
x
str(x)
```
```{r}
x <- list("a" = 2.5, "b" = TRUE, "c" = 1:3)
x
str(x)
```

#### Indexing list
Different from vector and matrix, for list, you need to double-layer square 
brackets, either by numeric index or name. Alternatively, you can also use `$` 
symbol.
```{r}
x[[3]]

x[["c"]]

x$c
```

### Data Frame
* A special type of List: A list of vectors with the same length.
* Widely used as a rectangular data with flexible data type (like Excel)

```{r}
df <- data.frame("SN" = 1:2, "Age" = c(21,15), 
                 "Name" = c("John","Dora"))
df
```

```{r}
df$Age[2]
```

```{r}
# View(df)
```

### Factor vs vector
Factor is a data structure used for fields that takes only predefined, finite 
number of values (categorical data)
```{r}
x = c("single", "married", "married", "single") # vector
x

class(x)
typeof(x)
```

```{r}
y = factor(c("single", "married", "married", "single"))
y

class(y)
typeof(y)
```

#### Change the order of factor levels
```{r}
z <- factor(c("single", "married", "married", "single") , levels=c("single", "married"))
z
```

#### Smaller memory as categorical data type
```{r}
x = c("single", "married", "married", "single") # vector
y = factor(c("single", "married", "married", "single"))

object.size(rep(x, 1000))
object.size(rep(y, 1000))
```


## Read and write files (tables)
Data is available in the data folder on the 
[github repository](https://github.com/StatBiomed/BMDatSci-book/blob/main/notebooks/chapter1-R)

### Read file
See full manuals: `help(read.table)` or `?read.table`
```{r}
help("read.table")

?read.table
```

```{r}
df = read.table("./SRP029880.colData.tsv", sep="\t")
df
```
### Write file

```{r}
df$frozen <- c(1, 1, 0, 0, 0, 1, 1, 0, 0, 0)
write.table(df, "./SRP029880.colData.add_frozen.tsv", sep="\t", quote=FALSE)
```

## Functions and Packages
For example `mean()` is a function here and it is from the `base` package
```{r}
x <- 4:10
mean(x)
base::mean(x)
```

### Install packages
It depends on where the package is stored. Please refers to the documentation 
of the specific package you want to install and use.

1. CRAN  (the Comprehensive R Archive Network): main platform
    - For example: `install.packages("ggplot2")`

2. Bioconductor: primarily for biology related packages
    - For example: `BiocManager::install("DESeq2")`

```{r}
#install.packages("ggplot2")
```

## Plotting
### datasets
Using a built-in dataset for illustration: iris (4 flower features in 3plants)
```{r}
head(iris)
```
```{r}
summary(iris)
```

### Basic plotting
#### Histogram
```{r fig1-hist}
hist(iris$Sepal.Length)
```

#### Scatter plot
```{r fig1-scatter}
plot(x=iris$Sepal.Length, y=iris$Sepal.Width)
```

#### boxplot
```{r fig1-box}
x1 <- iris$Sepal.Length[iris$Species == "setosa"]
x2 <- iris$Sepal.Length[iris$Species == "versicolor"]
x3 <- iris$Sepal.Length[iris$Species == "virginica"]

boxplot(x1, x2, x3)
```

### ggplot2
See more instructions: http://www.sthda.com/english/wiki/ggplot2-essentials

#### Install package
```{r}
if (!requireNamespace("ggplot2", quietly = TRUE))
    install.packages("ggplot2")
```

#### Load package
```{r}
library(ggplot2)
```

#### Histogram
```{r fig1-gghist}
ggplot(iris, aes(x=Sepal.Length)) + 
  geom_histogram(bins = 8)
```

#### Scatter plot
```{r fig1-ggpoint}
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width)) + 
    geom_point()
```

#### Box plot
```{r fig1-ggbox}
ggplot(iris, aes(x=Species, y=Sepal.Length)) + 
    geom_boxplot()
```

## Scientific and statistical computating
### Orders of operators
See lecture slides.

If you are not sure about a certain ordering, use brackets!

```{r}
5 * 2 > 4
```

```{r}
5 * (2 > 4)
```

### Functions for statistics
More theroy and practice to come in next session

### Correlation
```{r}
cor(iris$Sepal.Length, iris$Petal.Length)
```
```{r}
cor.test(iris$Sepal.Length, iris$Petal.Length)
```

### Hypothesis testing (t test)
```{r}
x1 <- iris$Sepal.Length[iris$Species == "setosa"]
x2 <- iris$Sepal.Length[iris$Species == "versicolor"]
x3 <- iris$Sepal.Length[iris$Species == "virginica"]

t.test(x2, x3)
```

### Regression
```{r}
fit <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data=iris)
summary(fit) # show results
```

This means the fitted regression is: 
`Sepal.Length ~ 1.856 + 0.65*Sepal.Width + 0.709*Petal.Length - 0.556*Petal.Width`

We can check how good the regression is by plotting it out
```{r}
y_pred <- fit$coefficients[1] + 
  fit$coefficients[2] * iris$Sepal.Width +
  fit$coefficients[3] * iris$Petal.Length +
  fit$coefficients[4] * iris$Petal.Width

cor(iris$Sepal.Length, y_pred)
```


```{r fig1-reg}
plot(iris$Sepal.Length, y_pred)
```


## Resource links

- https://www.datamentor.io/r-programming/
- https://www.geeksforgeeks.org/r-data-types/
- https://data-flair.training/blogs/r-data-types/
- http://www.sthda.com/english/wiki/ggplot2-essentials


