(ref:introsepart) Data Science Foundations

```{r echo=FALSE, results="asis", purl=FALSE}
cat("# (PART) Data Science Foundations {-} ")
```


# Introduction to R programming {#introR}

This notebook collects the scripts used for teaching in BIOF1001 for 
Introduction to R (1 hour teaching). You can get this Rmd file 
[here](https://github.com/StatBiomed/BMDatSci-book/blob/main/01-introR.Rmd)
(right click and "save link as" to download).

## Data types
### nemeric (or double)
```{r}
x <- c(1.0, 2.0, 5.0, 7.0)
x
class(x)
typeof(x)
```

### integer
```{r}
y <- c(1L, 2L, 5L, 7L)
y
typeof(y)
```

### logical
```{r}
z <- c(TRUE, TRUE, TRUE, FALSE)
z
typeof(z)
```

### character
```{r}
w <- c("aa", "bb", "5", "7")
w
typeof(w)
```

### Memeory usage
As can be seen below, integer and logical uses only half of the memory of double (numeric) and character
```{r}
x <- c(1.0, 2.0, 5.0, 7.0)
y <- c(1L, 2L, 5L, 7L)
z <- c(TRUE, TRUE, TRUE, FALSE)
w <- c("aa", "bb", "5", "7")

object.size(x)
```
```{r}
object.size(rep(x, 1000))
object.size(rep(y, 1000))
object.size(rep(z, 1000))
object.size(rep(w, 1000))
```

## Data structures
### Vector
```{r}
x <- c(1, 2, 5, 7)
x
typeof(x)

x <- rep(3, 5)
x
typeof(x)

x <- 1:12 # integer
x
typeof(x)
```
### Matrix
```{r}
A <- matrix(1:12, nrow=3)
A

B <- matrix(1:12, nrow=3, byrow=TRUE)
B

colnames(A) <- c("C1","C2","C3","C4")
rownames(A) <- c("R1","R2","R3")
A
```

#### Index Vector and Matrix
N.B., when using logical as index, be careful if the index length is different from the vector
```{r}
x <- 1:12

x[3]
x[2:5]
x[c(2, 5, 6)]

x[c(TRUE, FALSE, FALSE, TRUE)] 
```

Now, index matrix. Very similar as vector, but it has both row and column.
```{r}
A <- matrix(1:12, nrow=3)
colnames(A) <- c("C1","C2","C3","C4")
rownames(A) <- c("R1","R2","R3")

A[1, 2]
A[1, "C2"]

A[1, c(2, 3)]
A[1:2, c(2, 3)]
```
Single row or column matrix will become a vector, unless using `drop=FALSE`
```{r}
A[1, 2:4]
dim(A[1, 2:4])

A[1, 2:4, drop=FALSE]
dim(A[1, 2:4, drop=FALSE])
```
#### Modify values
```{r}
A[1, 2:4] <- c(-3, -5, 20)
A
```

### List
A list of any data structure: value, vector, matrix, etc
```{r}
x <- list(2.5, TRUE, 1:3)
x
str(x)
```
```{r}
x <- list("a" = 2.5, "b" = TRUE, "c" = 1:3)
x
str(x)
```

#### Indexing list
Different from vector and matrix, for list, you need to double-layer square brackets, either by numeric index or name. Alternatively, you can also use `$` symbol.
```{r}
x[[3]]

x[["c"]]

x$c
```

### Data Frame
* A special type of List: A list of vectors with the same length.
* Widely used as a rectangular data with flexible data type (like Excel)

```{r}
df <- data.frame("SN" = 1:2, "Age" = c(21,15), 
                 "Name" = c("John","Dora"))
df
```

```{r}
df$Age[2]
```

```{r}
# View(df)
```

### Factor vs vector
Factor is a data structure used for fields that takes only predefined, finite number of values (categorical data)
```{r}
x = c("single", "married", "married", "single") # vector
x

class(x)
typeof(x)
```

```{r}
y = factor(c("single", "married", "married", "single"))
y

class(y)
typeof(y)
```

#### Change the order of factor levels
```{r}
z <- factor(c("single", "married", "married", "single") , levels=c("single", "married"))
z
```

#### Smaller memory as categorical data type
```{r}
x = c("single", "married", "married", "single") # vector
y = factor(c("single", "married", "married", "single"))

object.size(rep(x, 1000))
object.size(rep(y, 1000))
```


## Read and write files (tables)
Data is available in the data folder on the 
[github repository](https://github.com/StatBiomed/BMDatSci-book/blob/main/data/intro)

### Read file
See full manuals: `help(read.table)` or `?read.table`
```{r}
help("read.table")

?read.table
```

```{r}
df = read.table("./data/intro/SRP029880.colData.tsv", sep="\t")
df
```
### Write file

```{r}
df$frozen <- c(1, 1, 0, 0, 0, 1, 1, 0, 0, 0)
write.table(df, "./data/outs/SRP029880.colData.add_frozen.tsv", sep="\t", quote=FALSE)
```

## Functions and Packages
For example `mean()` is a function here and it is from the `base` package
```{r}
x <- 4:10
mean(x)
base::mean(x)
```

### Install packages
It depends on where the package is stored. Please refers to the documentation of the specific package you want to install and use.

1. CRAN  (the Comprehensive R Archive Network): main platform
    - For example: `install.packages("ggplot2")`

2. Bioconductor: primarily for biology related packages
    - For example: `BiocManager::install("DESeq2")`

```{r}
#install.packages("ggplot2")
```

## Plotting
### datasets
Using a build-in dataset for illustration: iris (4 flower features in 3plants)
```{r}
head(iris)
```
```{r}
summary(iris)
```

### Basic plotting
#### Histogram
```{r}
hist(iris$Sepal.Length)
```

#### Scatter plot
```{r}
plot(x=iris$Sepal.Length, y=iris$Sepal.Width)
```

#### boxplot
```{r}
x1 <- iris$Sepal.Length[iris$Species == "setosa"]
x2 <- iris$Sepal.Length[iris$Species == "versicolor"]
x3 <- iris$Sepal.Length[iris$Species == "virginica"]

boxplot(x1, x2, x3)
```

### ggplot2
#### Install package
```{r}
if (!requireNamespace("ggplot2", quietly = TRUE))
    install.packages("ggplot2")
```

#### Load package
```{r}
library(ggplot2)
```

#### Histogram
```{r}
ggplot(iris, aes(x=Sepal.Length)) + 
  geom_histogram(bins = 8)
```

#### Scatter plot
```{r}
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width)) + 
    geom_point()
```

#### Box plot
```{r}
ggplot(iris, aes(x=Species, y=Sepal.Length)) + 
    geom_boxplot()
```

## Scientific and statistical computating
### Orders of operators
See lecture slides.

If you are not sure about a certain ordering, use brackets!

```{r}
5 * 2 > 4
```

```{r}
5 * (2 > 4)
```

### Functions for statistics
More theroy and practice to come in next session
### Correlation
```{r}
cor(iris$Sepal.Length, iris$Petal.Length)
```
```{r}
cor.test(iris$Sepal.Length, iris$Petal.Length)
```
### Hypothesis testing (t test)
```{r}
x1 <- iris$Sepal.Length[iris$Species == "setosa"]
x2 <- iris$Sepal.Length[iris$Species == "versicolor"]
x3 <- iris$Sepal.Length[iris$Species == "virginica"]

t.test(x2, x3)
```
### Regression
```{r}
fit <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width, data=iris)
summary(fit) # show results
```

This means the fitted regression is: 
`Sepal.Length ~ 1.856 + 0.65*Sepal.Width + 0.709*Petal.Length - 0.556*Petal.Width`

We can check how good the regression is by plotting it out
```{r}
y_pred <- fit$coefficients[1] + 
  fit$coefficients[2] * iris$Sepal.Width +
  fit$coefficients[3] * iris$Petal.Length +
  fit$coefficients[4] * iris$Petal.Width

cor(iris$Sepal.Length, y_pred)
```


```{r}
plot(iris$Sepal.Length, y_pred)
```



