

## Diagnostic metrics

Apart from the $R^2$ statistic, there are other statistics and parameters that 
you need to look at in order to determine if the model is efficient. We will 
discuss some commonly used statistics – Residual Standard Errors, $p$-values, 
and $F$-statistics, in the form of both simple linear regression and multiple
regression.

### Residual Standard Errors

Residual Standard Errors (RSE) is a common statistic used to calculate the 
accuracy of values predicted by 
a model. It is an estimate of the variance of the error term, `res`. For a 
simple linear regression model, RSE is defined as: 
$$  RSE^2 = \frac{SSE}{n-2} = \frac1{n-2} \sum_{i=1}^n  \Bigl(\text{yact}_i - \text{ypred}_i \Bigr)^2.
$$

In general for multiple regression, 

$$  RSE^2 = \frac{SSE}{n-p-1} = \frac1{n-p-1} \sum_{i=1}^n  \Bigl(\text{yact}_i - \text{ypred}_i \Bigr)^2.
$$
 
where $p$ is the number of predictor variables in a model where we have more 
than one predictor variables. The $p$ predictors do not include intercept.

 
A **multiple linear regression** model is a linear regression model with 
multiple predictors, written as  
$$  Y_e = \alpha +\beta_1 * X_1 +\cdots +\beta_p X_p.
$$

As you see, the parameters and predictors are subscripted from 1 up to the 
number of predictors $p$. 

In multiple regression, the value of RSE generally decreases as we add 
variables that are more significant predictors of the output variable.

Using our simulated data from the previous steps, the following code snippet 
shows how the RSE for a model can be calculated:

```{r}
set.seed(0)

## Generate data
X = 2.5 * rnorm(100) + 1.5   # Array of 100 values with mean = 1.5, stddev = 2.5
res = 0.5 * rnorm(100)       # Generate 100 residual terms
yact = 2 + 0.3 * X + res     # Actual values of Y

## Create dataframe to store our X, ypred, and yact values
df = data.frame('X' = X, 'yact' = yact)

## Calculate the mean of X and Y
xmean = mean(X)
ymean = mean(yact)

## Calculate the terms needed for the numator and denominator of beta
df['xycov'] = (df['X'] - xmean) * (df['yact'] - ymean)
df['xvar'] = (df['X'] - xmean)**2

## Calculate beta and alpha
beta = sum(df['xycov']) / sum(df['xvar'])
alpha = ymean - (beta * xmean)
print(paste('alpha =', alpha, ';', 'beta =', beta))

## Store predictions as in previous step
df['ypred'] = alpha + beta * df['X']

## Show first five rows of dataframe
head(df)
```


```{r}
## Calculate SSE
df['SSE'] = (df['yact'] - df['ypred'])**2
SSE = sum(df['SSE'])

## Calculate RSE
RSE = sqrt(SSE / 98)   # n = 100
print(paste('RSE =', RSE))
```

The value of `RSE` comes out to be 0.48. 

As you might have guessed, the smaller the residual standard errors, the better 
the model is. 

The benchmark to compare this to is the mean of the actual values, `yact`. As 
shown previously, this value is `ymean = 2.54`. In plain English, this means we 
observe an error of 0.48 over 2.44 - approximately 19.69%.


```{r}
error = RSE / ymean
print(paste('Mean Y =', ymean))
print(paste('Error =', error))
```

### p-values

The calculation of $\alpha$  and $\beta$ are estimates, not exact calculations. 
Whether their values are significant or not needs to be tested using a 
**hypothesis test**.

In the equation, $Y = \alpha + \beta X$, if we set $\beta=0$,  there will be no 
relation between $Y$ and $X$. Therefore, the hypothesis tests whether the value 
of $\beta$ is non-zero or not.

$$\begin{align*} \text{Null hypothesis}~  H_0~:~  \beta=0, & \quad \text{versus} \\
\text{Alternative hypothesis}~ H_1~:~ \beta\ne 0.&  \end{align*}  $$
 

Whenever a regression task is performed and $\beta$ is calculated, there will 
be an accompanying **p-value** corresponding to this hypothesis test. We will 
not go through how this is calculated in this course (you can learn more
[here](https://www.dummies.com/education/math/statistics/how-to-determine-a-p-value-when-testing-a-null-hypothesis/)), 
since it is calculated automatically by ready-made methods in R.

If the p-value is less than a chosen **significance level** (e.g. 0.05) then 
the null hypothesis that $\beta = 0$ is rejected and $\beta$ is said to be 
<b>significant and non-zero</b>.

In the case of multiple linear regression, the p-value associated with each 
$\beta_k$   can be used to weed out insignificant predictors from the model. 
The higher the p-value for $\beta_k$, the less significant $X_k$  is to the 
model.

### F-statistics

#### The Global F-Test

In a multiple regression model, testing the significance of individual variables 
via $p$-values is not sufficient. It is also necessary to determine whether the 
predictors are significant **as a group**. This is achieved by testing the 
following hypotheses:

$$
\begin{aligned} 
H_0: & \quad \beta_1 = \beta_2 = \dots = \beta_p = 0 \\
H_1: & \quad \text{At least one } \beta_k \neq 0 
\end{aligned}
$$

Conceptually, the null hypothesis ($H_0$) suggests that the entire set of 
features has no predictive power (i.e., the model is no better than the mean), 
whereas the alternative hypothesis ($H_1$) suggests that at least one feature 
is useful.

**The F-Statistic Formula**
The statistic used to test this is the **F-statistic** (or Fisher statistic). 
It is defined as the ratio of the variance explained by the model to the 
unexplained variance (noise), each adjusted by their respective degrees of 
freedom (DF):

$$ F\text{-statistic} = \frac{\text{Explained Variance} / \text{Model DF}}{\text{Unexplained Variance} / \text{Error DF}} $$

In the context of linear regression, this is expressed explicitly as:

$$ F = \frac{SSR / p}{SSE / (n - p - 1)} $$

Where:

*   **$n$**: The number of observations (sample points).
*   **$p$**: The number of predictor variables (not including intercept).
*   **$SSR$**: Sum of Squares due to Regression ($SST - SSE$).
*   **$SSE$**: Sum of Squared Errors (Residual Sum of Squares).

**Interpretation**
An associated $p$-value is calculated for this F-statistic based on the 
$F(p, n-p-1)$ distribution. If this $p$-value is smaller than the chosen 
significance level (e.g., $\alpha = 0.05$), we reject the null hypothesis and 
conclude that the model is statistically significant, meaning some of the 
features are useful.


#### Why is the F-statistic necessary?

It is important to look at the F-statistic rather than just individual 
$p$-values for two main reasons:

1.  **Interdependence of Predictors**: Individual $p$-values assess the 
  relationship between a single predictor and the outcome, assuming other 
  variables are held constant. If predictors are highly correlated 
  (multicollinearity), their individual significance may be obscured. The 
  F-statistic looks at the collective effect.
2.  **The Multiple Comparisons Problem**: If a model contains a very large 
  number of predictors, there is a high probability that at least one predictor 
  will appear significant by random chance (a Type I error), even if all 
  $\beta_i$ are actually zero. The F-statistic protects against this "false 
  discovery" by providing a single test for the entire model.


#### The Partial F-Test (Testing Sub-groups)

One may also use the F-statistic to assess a specific sub-group of features 
(for example, the last $q$ features added to a model). This helps determine if 
adding those specific variables significantly improves the model.

**Hypotheses:**

*   **$H_0$**: $\beta_{p-q+1} = \beta_{p-q+2} = \dots = \beta_{p} = 0$
*   **$H_1$**: At least one of these $q$ coefficients is non-zero.

Let $SSE_{full}$ be the error sum of squares for the model with all $p$ 
features, and $SSE_{sub}$ be the error sum of squares for the reduced model 
(after dropping the $q$ features). The F-statistic is:

$$ F = \frac{(SSE_{sub} - SSE_{full}) / q}{SSE_{full} / (n - p - 1)} $$

**Connection to the Global Test**
The default "global" F-statistic is a special case of this formula. When you 
drop *all* features ($q = p$), the reduced model only contains the intercept, 
making $SSE_{sub}$ equal to the **Total Sum of Squares (SST)**.
