

## Diagnostic metrics

Apart from the $R^2$ statistic, there are other statistics and parameters that 
you need to look at in order to determine if the model is efficient. We will 
discuss some commonly used statistics – Residual Standard Errors, $p$-values, 
and $F$-statistics.

### Residual Standard Errors

Residual Standard Errors (RSE) is a common statistic used to calculate the 
accuracy of values predicted by 
a model. It is an estimate of the variance of the error term, `res`. For a 
simple linear regression model, RSE is defined as: 
$$  RSE^2 = \frac{SSE}{n-2} = \frac1{n-2} \sum_{i=1}^n  \Bigl(\text{yact}_i - \text{ypred}_i \Bigr)^2.
$$

In general, 

$$  RSE^2 = \frac{SSE}{n-p-1} = \frac1{n-p-1} \sum_{i=1}^n  \Bigl(\text{yact}_i - \text{ypred}_i \Bigr)^2.
$$
 
where $p$ is the number of predictor variables in a model where we have more 
than one predictor variables.

 
A **multiple linear regression** model is a linear regression model with 
multiple predictors, written as  
$$  Y_e = \alpha +\beta_1 * X_1 +\cdots +\beta_p X_p.
$$

As you see, the parameters and predictors are subscripted from 1 up to the 
number of predictors $p$. 

In multiple regression, the value of RSE generally decreases as we add 
variables that are more significant predictors of the output variable.

Using our simulated data from the previous steps, the following code snippet 
shows how the RSE for a model can be calculated:

```{r}
set.seed(0)

## Generate data
X = 2.5 * rnorm(100) + 1.5   # Array of 100 values with mean = 1.5, stddev = 2.5
res = 0.5 * rnorm(100)       # Generate 100 residual terms
yact = 2 + 0.3 * X + res     # Actual values of Y

## Create dataframe to store our X, ypred, and yact values
df = data.frame('X' = X, 'yact' = yact)

## Calculate the mean of X and Y
xmean = mean(X)
ymean = mean(yact)

## Calculate the terms needed for the numator and denominator of beta
df['xycov'] = (df['X'] - xmean) * (df['yact'] - ymean)
df['xvar'] = (df['X'] - xmean)**2

## Calculate beta and alpha
beta = sum(df['xycov']) / sum(df['xvar'])
alpha = ymean - (beta * xmean)
print(paste('alpha =', alpha, ';', 'beta =', beta))

## Store predictions as in previous step
df['ypred'] = alpha + beta * df['X']

## Show first five rows of dataframe
head(df)
```


```{r}
## Calculate SSE
df['SSE'] = (df['yact'] - df['ypred'])**2
SSE = sum(df['SSE'])

## Calculate RSE
RSE = sqrt(SSE / 98)   # n = 100
print(paste('RSE =', RSE))
```

The value of `RSE` comes out to be 0.48. 

As you might have guessed, the smaller the residual standard errors, the better 
the model is. 

The benchmark to compare this to is the mean of the actual values, `yact`. As 
shown previously, this value is `ymean = 2.54`. In plain English, this means we 
observe an error of 0.48 over 2.44 - approximately 19.69%.


```{r}
error = RSE / ymean
print(paste('Mean Y =', ymean))
print(paste('Error =', error))
```

### p-values

The calculation of $\alpha$  and $\beta$ are estimates, not exact calculations. 
Whether their values are significant or not needs to be tested using a 
**hypothesis test**.

In the equation, $Y = \alpha + \beta X$, if we set $\beta=0$,  there will be no 
relation between $Y$ and $X$. Therefore, the hypothesis tests whether the value 
of $\beta$ is non-zero or not.

$$\begin{align*} \text{Null hypothesis}~  H_0~:~  \beta=0, & \quad \text{versus} \\
\text{Alternative hypothesis}~ H_1~:~ \beta\ne 0.&  \end{align*}  $$
 

Whenever a regression task is performed and $\beta$ is calculated, there will 
be an accompanying **p-value** corresponding to this hypothesis test. We will 
not go through how this is calculated in this course (you can learn more
[here](https://www.dummies.com/education/math/statistics/how-to-determine-a-p-value-when-testing-a-null-hypothesis/)), 
since it is calculated automatically by ready-made methods in R.

If the p-value is less than a chosen **significance level** (e.g. 0.05) then 
the null hypothesis that $\beta = 0$ is rejected and $\beta$ is said to be 
<b>significant and non-zero</b>.

In the case of multiple linear regression, the p-value associated with each 
$\beta_k$   can be used to weed out insignificant predictors from the model. 
The higher the p-value for $\beta_k$, the less significant $X_k$  is to the 
model.

### F-statistics

In a multiple regression model, apart from testing the significance of 
individual variables by checking the p-values, it is also necessary to check 
whether, as a group all the predictors are significant. This can be done using 
the following hypothesis:

$$\begin{align*} \text{Null hypothesis}~  H_0~:~ & \beta_1=\beta_2=\cdots=\beta_p=0, \quad \text{versus} \\
\text{Alternative hypothesis}~ H_1~:~& \text{at least one of the} ~\beta_k's ~ \text{is non zero}. \end{align*}  $$
 

The statistic that is used to test this hypothesis is called the **F-statistic** 
and is defined as follows:

$$  F\text{-statistic} = \text{Fisher statistic}=  \frac{ SSR/(p-1)}{ SSE/(n-p)}
$$

where $n$ = number of rows (sample points) in the dataset, $p$ = number of 
predictor variables in the model, and $SSR = SST - SSE$ is the sum of squares 
due to regression.

There is a $p$-value that is associated with this $F$-statistic. If the 
$p$-value is smaller than the chosen significance level, the null hypothesis 
can be rejected.

It is important to look at the F-statistic because:

- p-values are about individual relationships between predictors and the outcome
variable. However, one predictor's relationship with the output might be 
impacted by the presence of other variables.
- When the number of predictors in the model is very large and all the 
$\beta_i$ are very close to zero, the individual p-values associated with the 
predictors might give very small values so we might incorrectly conclude that 
there is a relationship between the predictors and the outcome.

