
## Multiple testing

Hypothetical null distribution. 
Feel feel to try any null distribution, examples below

```{r}
## Example null distributions
# t, normal or anything. we use chi-squared distribution as an example

x_random = rchisq(n=1000, df=3)
any_null_dist = dchisq(x_random, df=3)

pvals_null = 1 - pchisq(x_random, df=3)
```


### Null distribution (of test statistic)


```{r fig4-stat-null, out.width = '60%'}
# Null distribution of test statistic
hist(x_random)
```

### Null distribution of p value

```{r fig4-pval-null, out.width = '60%'}
# Null distribution of test statistic
hist(pvals_null)
```

### Minimal p values in 10 tests

```{r}
# We use matrix to group 100 trials into a column
# We then use apply() to calculate min value for each column

pval_null_mtx = matrix(pvals_null, nrow=10)

p_min_in10 = apply(pval_null_mtx, MARGIN=2, FUN=min)
```

```{r fig4-pmin-null, out.width = '60%'}
hist(p_min_in10)
```

```{r}
print(paste('Proportion of tests with min(p) < 0.05:', mean(p_min_in10 < 0.05)))
```

## Explore power and sample size (optional) { power }

1. Make a simulation of score: group A and B
2. B follows normal(mean=0, std=1); A follows normal(mean=0.1, std=1)
3. Generate 100 samples for each group, and do a t test, is difference significant? Please use set.seed(0) beforehand.
4. Try 3) again but general 3,00 samples this time, later 1,000 samples. What 
   do you find? Think the relation between power and sample size.

```{r}
set.seed(0)

n_sample = 100 # change this value to 1000 and 10000
xB = rnorm(n_sample)
xA = rnorm(n_sample, mean=0.1)
t.test(xA, xB, var.equal = TRUE)
```


