

## Visualise logistic and logit functions

In this chapter, we will focus on logistic regression for classification. 
Let's first look at what logistic and logit function look like.

### Logistic function

Let's write our first function `logestic()` as follows.
```{r}
# Write your first function
logistic <- function(y) {
  exp(y) / (1 + exp(y))
}

# Try it with different values:
logistic(0.1)
logistic(c(-3, -2, 0.5, 3, 5))
```

This is the equivalent to the built-in `plogis()` function in the `stat` 
package for the 
[logistic distribution](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/Logistic):

```{r}
plogis(0.1)
plogis(c(-3, -2, 0.5, 3, 5))
```


### Logit function
Now, let look at the logistic's inverse function `logit()`, and let's define it
manually. **Note**, this function only support input between 0 and 1.
```{r}
# Write your first function
logit <- function(x) {
  log(x / (1 - x))
}

# Try it with different values:
logit(0.4)
logit(c(0.2, 0.3, 0.5, 0.7, 0.9))
logit(c(-1, 2, 0.4))
```

Again, the built-in `stat` package's logistic distribution has an equivalent 
function `qlogis()`, though with a different name.

```{r}
qlogis(0.4)
qlogis(c(0.2, 0.3, 0.5, 0.7, 0.9))
qlogis(c(-1, 2, 0.4))
```

### Visualise the distribution
Logisitc function
```{r fig3-logistic, out.width = '60%'}
# You can use seq() function to generate a vector
# Check how to use it by help(seq) or ?seq
x = seq(-7, 7, 0.3)
df = data.frame('x'=x, 'logistic'=plogis(x))

# You can plot by plot function
# plot(x=df$x, y=df$logistic, type='o')

# Or ggplot2
library(ggplot2)
ggplot(df, aes(x=x, y=logistic)) +
  geom_point() + geom_line()
```


Logit function
```{r fig3-logit, out.width = '60%'}
x = seq(0.001, 0.999, 0.01)
df = data.frame('x'=x, 'logit'=qlogis(x))

ggplot(df, aes(x=x, y=logit)) +
  geom_point() + geom_line()
```

