
## Case study 2: Cancer Epidemiology { #cancer-case2 }

*by Prof. Jason Wong*

*Date: 31-10-2024*

[The RMarkdown notebook to run your own code can be downloaded here](https://github.com/StatBiomed/BMDS-book/raw/main/notebooks/module5-epidemi/M5-epidemiology.Rmd) [right click -> save link as]

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

### Scenario

You are a grants officer working for the Hong Kong Health Bureau. The HK government has recently announced new special funding in cancer research to be administered by the Bureau. You are tasked with coming up with a proposal for distribution of funding to specific cancer types that are in most need of research.

[Discussion points:]{.underline}

-   What is important to consider when selecting a cancer type in need of research?

-   What type of data is required?

### Hong Kong population

You are aware that, generally, cancer is a disease that affects the elderly more than the young. You decide first to take a closer look at the structure of the population of Hong Kong.

[Historical population of Hong Kong](https://www.censtatd.gov.hk/en/web_table.html?id=1A) can be obtained from the Census and Statistics Department.

An abridged version of the full historical population of Hong Kong is provided here, containing the population of Hong Kong from 1965, 1975, 1985, 1995, 2005, 2015 and 2022 categorised by sex and age (0-19, 20-44, 45-64, 65+).

[Discussion points:]{.underline}

-   What is the trend in Hong Kong's population over the past \~60 years?

-   What is the best way to visualise this data?

#### Download population data

```{r loadpopulation}
HKPop<- read.table("https://github.com/StatBiomed/BMDS-book/raw/main/notebooks/module5-epidemi/HK_population_1965-2022.txt", 
    sep = "\t", header = TRUE, stringsAsFactor=FALSE, check.names = FALSE)
HKPop
```

#### Format data for plotting

Here we convert the original dataframe into simplified format for ggplot2.

```{r format data}

male<-data.frame(year = as.numeric(colnames(HKPop[1,3:9])),
                 `0-19`  = as.numeric(HKPop[1,3:9]),
                 `20-44` = as.numeric(HKPop[2,3:9]),
                 `45-64` = as.numeric(HKPop[3,3:9]),
                 `65+` = as.numeric(HKPop[4,3:9]),
                 check.names = FALSE)



female<-data.frame(year = as.numeric(colnames(HKPop[1,3:9])),
                 `0-19`  = as.numeric(HKPop[5,3:9]),
                 `20-44` = as.numeric(HKPop[6,3:9]),
                 `45-64` = as.numeric(HKPop[7,3:9]),
                 `65+` = as.numeric(HKPop[8,3:9]),
                 check.names = FALSE)

if (!require("tidyverse")) install.packages("tidyverse")
male<-as_tibble(male) %>% select(year,`0-19`,`20-44`,`45-64`,`65+`) %>% gather (key="age",value="population ('000s)",-year)
female<-as_tibble(female) %>% select(year,`0-19`,`20-44`,`45-64`,`65+`) %>% gather (key="age",value="population ('000s)",-year)
male
female
```

#### Plotting population data

Uses ggplot2 and gridExtra to make line plot of male and female population data side-by-side.

```{r plot data}

if (!require("ggplot2")) install.packages("gglot2")
if (!require("gridExtra")) install.packages("gridExtra")
library(ggplot2)
library(gridExtra)
#Import the necessary packages and libraries
pmale<-ggplot(male,aes(x=year,y=`population ('000s)`,group=age))+
  geom_line(aes(color=age))+
  geom_point(aes(color=age))+
  scale_color_brewer(palette="Spectral")+
  theme_classic()+
  ylim(0,1600)+
  theme(legend.position="none")+
  scale_x_continuous(breaks = seq(1965, 2022, by = 10))+
  ggtitle("male")
pfemale<-ggplot(female,aes(x=year,y=`population ('000s)`,group=age))+
  geom_line(aes(color=age))+
  geom_point(aes(color=age))+
  scale_color_brewer(palette="Spectral")+
  theme_classic()+
  ylim(0,1600)+
  theme(legend.position="right",axis.title.y = element_blank())+
  scale_x_continuous(breaks = seq(1965, 2022, by = 10))+
  ggtitle("female")
grid.arrange(pmale,pfemale,ncol=2,widths=c(3,3.75))
```

### Cancer registry data

It is clear that Hong Kong has an aging population, thus cancer incidence would also likely increase. To examine cancer incidence and mortality in Hong Kong you obtain data from the [Hong Kong Cancer Registry](https://www3.ha.org.hk/cancereg/), which is maintained by the Hospital Authority.

Cancer incidence data was summarised for the last three decades (1990-1999, 2000-2009 and 2010-2019).

[Discussion points:]{.underline}

-   Has incidence been increasing for most cancers? How about mortality?

-   How has cancer risk and mortality rate changed in the past 3 decades?

-   Has the incidence-to-mortality ratio been decreasing generally? Is it statistically significant?

-   Which cancer type has the highest incidence in children (0-19) when compared with the elderly (65+). Is this statistically significantly different to incidence of children versus elderly cancers in general?

#### Download cancer registry data

```{r load cancer data}
HKCancer<- read.table("https://github.com/StatBiomed/BMDS-book/raw/main/notebooks/module5-epidemi/HK_cancer_incidence_mortality_1990-2020.txt",
                      sep = "\t", header = TRUE, stringsAsFactor=FALSE)
HKCancer
```

#### a. Visualise changes in incidence and mortality

```{r incidence, fig.height=20, fig.width=14}
# first plot incidence for each cancer type

#Import the necessary packages and libraries
if (!require("ggplot2")) install.packages("gglot2")
if (!require("gridExtra")) install.packages("gridExtra")
if (!require("ggpubr")) install.packages("ggpubr")
library(ggplot2)
library(gridExtra)
library(ggpubr)

HKCancer_inc <- HKCancer[HKCancer$Type=='incidence',]

p<-list()

for (i in 1:(ncol(HKCancer_inc)-4)){
  p[[i]]<-ggplot(HKCancer_inc,aes_string(fill=names(HKCancer_inc)[3],x=names(HKCancer_inc)[4],
                                         y=names(HKCancer_inc)[i+4],group=names(HKCancer_inc)[3]))+
       geom_bar(position="dodge",stat="identity")+
       facet_wrap(~Sex) +
       theme_classic()+
       scale_fill_brewer(palette="Spectral")+
       theme(legend.position="none")+
       theme(text = element_text(size = 10))+ 
       ggtitle(names(HKCancer_inc)[i+4])+
       ylab("Incidence")
}
do.call('grid.arrange',c(p,ncol=3,nrow=12))

```

```{r mortality, fig.height=20, fig.width=14}
# now plot mortality for each cancer type

#Import the necessary packages and libraries
if (!require("ggplot2")) install.packages("gglot2")
if (!require("gridExtra")) install.packages("gridExtra")
if (!require("ggpubr")) install.packages("ggpubr")
if (!require("dplyr")) install.packages("dplyr")
library(ggplot2)
library(gridExtra)
library(ggpubr)
library(dplyr)

HKCancer_mort <- HKCancer[HKCancer$Type=='mortality',]

p<-list()
#(ncol(HKCancer_inc)-4)
for (i in 1:(ncol(HKCancer_mort)-4)){
  p[[i]]<-ggplot(HKCancer_mort,aes_string(fill=names(HKCancer_mort)[3],x=names(HKCancer_mort)[4],
                                          y=names(HKCancer_mort)[i+4],group=names(HKCancer_mort)[3]))+
       geom_bar(position="dodge",stat="identity")+
       facet_wrap(~Sex) +
       theme_classic()+
       scale_fill_brewer(palette="Spectral")+
       theme(legend.position="none")+
       theme(text = element_text(size = 10))+ 
       ggtitle(names(HKCancer_mort)[i+4])+
       ylab("Mortality")
}
do.call('grid.arrange',c(p,ncol=3,nrow=12))

```

#### b. Calculate cancer risk and mortality rate

To calculate disease risk we need to calculated the number of new cases over the number of persons at risk over a specific time period. We have the incidence for each decade and can estimate the number of persons at risk based on the population in 1995, 2005 and 2015.

```{r cancer risk}
# cancer risk calculation

HKCancer_inc <- HKCancer[HKCancer$Type=='incidence',]
HKCancer_inc_risk <- HKCancer_inc[,1:5]

risk <- function(x,age,sex,year){
  if (is.integer(x)){
    pop_n <- HKPop %>% filter(Sex==sex & Age==age)
    if (year == "1990-1999"){ return (as.double(x)/as.double(pop_n$`1995`)) }
    else if (year == "2000-2009") {return (as.double(x)/as.double(pop_n$`2005`))}
    else { return (as.double(x)/as.double(pop_n$`2015`)) }
  }
  return (x)
}

for (i in 5:ncol(HKCancer_inc)){
  HKCancer_inc_risk[names(HKCancer_inc)[i]] <- mapply(risk,HKCancer_inc[,i],HKCancer_inc[,3],HKCancer_inc[,2],HKCancer_inc[,4])
}
```

```{r cancer risk visulisation, fig.height=20, fig.width=14}
#visualise cancer risk

p<-list()
for (i in 1:(ncol(HKCancer_inc_risk)-4)){
  p[[i]]<-ggplot(HKCancer_inc_risk,aes_string(fill=names(HKCancer_inc_risk)[3],x=names(HKCancer_inc_risk)[4],
                                              y=names(HKCancer_inc_risk)[i+4],group=names(HKCancer_inc_risk)[3]))+
       geom_bar(position="dodge",stat="identity")+
       facet_wrap(~Sex) +
       theme_classic()+
       scale_fill_brewer(palette="Spectral")+
       theme(legend.position="none")+
       theme(text = element_text(size = 10))+ 
       ggtitle(names(HKCancer_inc)[i+4])+
       ylab("incidence per 1000")
}
do.call('grid.arrange',c(p,ncol=3,nrow=12))

```

```{r mortality rate calculation}
# mortality rate calculation

HKCancer_mort <- HKCancer[HKCancer$Type=='mortality',]
HKCancer_mort_risk <- HKCancer_mort[,1:5]

risk <- function(x,age,sex,year){
  if (is.integer(x)){
    pop_n <- HKPop %>% filter(Sex==sex & Age==age)
    if (year == "1990-1999"){ return (as.double(x)/as.double(pop_n$`1995`)) }
    else if (year == "2000-2009") {return (as.double(x)/as.double(pop_n$`2005`))}
    else { return (as.double(x)/as.double(pop_n$`2015`)) }
  }
  return (x)
}

for (i in 5:ncol(HKCancer_mort)){
  HKCancer_mort_risk[names(HKCancer_mort)[i]] <- mapply(risk,HKCancer_mort[,i],HKCancer_mort[,3],HKCancer_mort[,2],HKCancer_mort[,4])
}

```

```{r mortality rate visulisation, fig.height=20, fig.width=14}
#visualise mortality rate

p<-list()
for (i in 1:(ncol(HKCancer_mort_risk)-4)){
  p[[i]]<-ggplot(HKCancer_mort_risk,aes_string(fill=names(HKCancer_mort_risk)[3],x=names(HKCancer_mort_risk)[4],
                                               y=names(HKCancer_mort_risk)[i+4],group=names(HKCancer_mort_risk)[3]))+
       geom_bar(position="dodge",stat="identity")+
       facet_wrap(~Sex) +
       theme_classic()+
       scale_fill_brewer(palette="Spectral")+
       theme(legend.position="none")+
       theme(text = element_text(size = 10))+ 
       ggtitle(names(HKCancer_inc)[i+4])+
       ylab("mortality per 1000")
}
do.call('grid.arrange',c(p,ncol=3,nrow=12))

```

#### c. Mortality-incidence ratio

Cancer research can be focused on improving cancer outcomes in a number of ways. For example cancer prevention research that seeks to reduce cancer incidence which would also ultimately reduce cancer mortality. Another area is cancer therapy which would not affect incidence but seeks to reduce mortality, or at least prolong survival. We don't go into survival analysis in this tutorial, but a way to get an idea whether treatment is improving by looking at the mortality-incidence ratio.

```{r mortality-incidence ratio calculation}

HKCancer_inc <- HKCancer[HKCancer$Type=='incidence',]
HKCancer_mort <- HKCancer[HKCancer$Type=='mortality',]
HKCancer_mort_inc <- HKCancer_mort[,1:5]

risk <- function(mort,inc){
  if (inc == 0){ return (0) }
  return (as.double(mort)/as.double(inc))
}

for (i in 5:ncol(HKCancer_mort)){
  HKCancer_mort_inc[names(HKCancer_mort)[i]] <- mapply(risk,HKCancer_mort[,i],HKCancer_inc[,i])
}

```

```{r mortality incidence ratio visulisation, fig.height=20, fig.width=14}
#visualise mortality incidence ratio

p<-list()
for (i in 1:(ncol(HKCancer_mort_inc)-4)){
  p[[i]]<-ggplot(HKCancer_mort_inc,aes_string(fill=names(HKCancer_mort_inc)[3],x=names(HKCancer_mort_inc)[4],
                                              y=names(HKCancer_mort_inc)[i+4],group=names(HKCancer_mort_inc)[3]))+
       geom_bar(position="dodge",stat="identity")+
       facet_wrap(~Sex) +
       theme_classic()+
       scale_fill_brewer(palette="Spectral")+
       theme(legend.position="none")+
       theme(text = element_text(size = 10))+ 
       ggtitle(names(HKCancer_inc)[i+4])+
       ylab("mortality-incidence ratio")
}
do.call('grid.arrange',c(p,ncol=3,nrow=12))

```

#### d. Paired t-test on mortality-incidence ratio change

In general, across the different cancer types is cancer treatment improving? We can use a paired t-test comparing the mortality-incidence ratio of cancers from the 1990-1999 period with the 2010-2019 period.

```{r paired t-test, fig.width=5}

#First sum up all incidence and mortality data for each cancer type across age and sex

HKCancer_inc_sum <- aggregate(HKCancer_inc[,-(1:4)],list(HKCancer_inc$Year),FUN=sum)
HKCancer_mort_sum <- aggregate(HKCancer_mort[,-(1:4)],list(HKCancer_mort$Year),FUN=sum)
HKCancer_mort_inc_year <- data.frame(Year=HKCancer_mort_sum[,1])

risk <- function(mort,inc){
  return (as.double(mort)/as.double(inc))
}

for (i in 2:ncol(HKCancer_mort_sum)){
  HKCancer_mort_inc_year[names(HKCancer_mort_sum)[i]] <- mapply(risk,HKCancer_mort_sum[,i],HKCancer_inc_sum[,i])
}

HKCancer_mort_inc_year_m <-data.frame(`1990-1999` = as.numeric(HKCancer_mort_inc_year[1,2:37]),
                                    `2000-2009` = as.numeric(HKCancer_mort_inc_year[2,2:37]),
                                    `2010-2019` = as.numeric(HKCancer_mort_inc_year[3,2:37]),
                                    check.names = FALSE)

HKCancer_mort_inc_year_m$Cancer <- names(HKCancer_mort_inc_year)[-1]

HKCancer_mort_inc_year_t<-as_tibble(HKCancer_mort_inc_year_m) %>% select(`Cancer`,`1990-1999`,`2000-2009`,`2010-2019`)%>% gather (key="Year",value="MIR",-Cancer)

plot<-ggplot(HKCancer_mort_inc_year_t,aes(x=Year,y=MIR, color=Year))+
       geom_boxplot(na.rm=T) +
       theme_classic()+
       scale_color_brewer(palette="Dark2")+
       geom_jitter(shape=16, position=position_jitter(0.2),na.rm=T)+
       ylab("mortality-incidence ratio")+
       ylim(0,1.2) + geom_signif(comparisons = list(c("1990-1999", "2010-2019")),
                                 map_signif_level=F, test= "t.test",test.args = list(paired = TRUE), na.rm = T, y_position = c(1.1, 1.3)) +
       geom_signif(comparisons = list(c("1990-1999", "2000-2009")), map_signif_level=F, test= "t.test",test.args = list(paired = TRUE), na.rm = T) +
       geom_signif(comparisons = list(c("2000-2009", "2010-2019")), map_signif_level=F, test= "t.test",test.args = list(paired = TRUE), na.rm = T)
plot

p<-list()
p[[1]]<-t.test(HKCancer_mort_inc_year_m$`1990-1999`,HKCancer_mort_inc_year_m$`2000-2009`,paired=TRUE,alternative_m = "two.sided")
p[[2]]<-t.test(HKCancer_mort_inc_year_m$`1990-1999`,HKCancer_mort_inc_year_m$`2010-2019`,paired=TRUE,alternative_m = "two.sided")
p[[3]]<-t.test(HKCancer_mort_inc_year_m$`2000-2009`,HKCancer_mort_inc_year_m$`2010-2019`,paired=TRUE,alternative_m = "two.sided")
p

```

#### e. Childhood versus elderly cancers

Although it is clear that the incidence of cancer is typically higher in the elderly, some cancers affect children as well. What cancer types disproportionate affect children? For each cancer type, compare the proportion of 0-19 versus 65+ incidence against the 0-19 versus 65+ incidence for all other cancer types.

```{r childhood versus elderly}
# Examine the proportion of childhood 

HKCancer_inc <- HKCancer[HKCancer$Type=='incidence',]
HKCancer_inc_age_sum <- aggregate(HKCancer_inc[,-(1:4)],list(HKCancer_inc$Age),FUN=sum)
HKCancer_inc_age_sum$Total<- rowSums(HKCancer_inc_age_sum[,-1])
HKCancer_inc_age_sum_csq <-  data.frame(cancer=names(HKCancer_inc_age_sum[,2:ncol(HKCancer_inc_age_sum)]))

pval <- list()
ratio <- list()
for (i in 2:(ncol(HKCancer_inc_age_sum))){
  val = Map('-',HKCancer_inc_age_sum$Total,HKCancer_inc_age_sum[,i])
  dat <- data.frame(cancer=HKCancer_inc_age_sum[c(1,4),i], other =c(val[[1]],val[[2]]))
  pval <- append(pval,fisher.test(dat)$p.val)
  ratio<- append(ratio,log(as.double(dat[1,1]+0.1)/as.double(dat[2,1]+0.1)))
}
HKCancer_inc_age_sum_csq$pval <- pval
HKCancer_inc_age_sum_csq$ratio <- ratio 

plot_child<-ggplot(HKCancer_inc_age_sum_csq[-37,],aes(x=reorder(cancer,-as.numeric(ratio)),y=as.numeric(ratio)),fill=cancer)+
       geom_bar(stat="identity",fill="red")+
       geom_hline(yintercept=-4.12132318942113, linetype="dashed", 
                color = "black", linewidth=0.5)+
       theme_classic()+
       scale_fill_brewer(palette="Spectral")+
       theme(legend.position="none")+
       #theme(text = element_text(size = 10))+
       theme(axis.title.x=element_blank())+
       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
       ylab("log((child+0.1)/(elderly+0.1)")
plot_child
HKCancer_inc_age_sum_csq
```

### Existing cancer funding and publication data

The Hong Kong government established in Health and Medical Research Fund (HMRF) in 2011 to specifically provide research funding for health and medical research in Hong Kong. Since 2016 over 370 projects in the category of Cancer has been funded for a total of \~\$400 M dollars. A list of all funded projects can be found on the Health Bureau [webpage](https://rfs1.healthbureau.gov.hk/search/#/fundedsearch/basicsearch?lang=en). You would like to use this data to see if there is any association between previous project funding and the epidemiology of cancers in Hong Kong.

We can also do a similar thing with publications and ask if the research publications in Hong Kong have been aligned with the incidence and mortality. We can obtain this data from PubMed using the following terms: [("Hong Kong"[Affiliation]) AND (neoplasms[MeSH Terms])](https://pubmed.ncbi.nlm.nih.gov/?term=%28%22Hong+Kong%22%5BAffiliation%5D%29+AND+%28neoplasms%5BMeSH+Terms%5D%29)

The data has been predownloaded as the Pubmed API via R is a bit slow.

[Discussion points:]{.underline}

-   Why has research publications increased dramatically in recent years? Is there something unusual with the dataset?
-   What are the main cancer types being researched in Hong Kong?
-   Is there any correlation between funding and cancer incidence and mortality?

#### Download HMRF grants and Pubmed data

```{r load HMRF}

HMRF<- read.delim("https://github.com/StatBiomed/BMDS-book/raw/main/notebooks/module5-epidemi/HMRF_cancer_grants.txt", sep = "\t", header = TRUE, stringsAsFactor=FALSE, check.names = TRUE)
#HMRF

pubmed<- read.delim("https://github.com/StatBiomed/BMDS-book/raw/main/notebooks/module5-epidemi/pubmed__neoplasm_hongkong.txt", sep = "\t", header = TRUE, stringsAsFactor=FALSE, check.names = TRUE)
#pubmed

hist(pubmed$Publication.Year)

```

#### Make word cloud for grants

<!-- ```{r wordcloud grants, fig.height=8, fig.width=10} -->
<!-- if (!require("ggwordcloud")) install.packages("ggwordcloud") -->
<!-- if (!require("tm")) install.packages("tm") -->
<!-- if (!require("dplyr")) install.packages("dplyr") -->
<!-- if (!require("RColorBrewer")) install.packages("RColorBrewer") -->
<!-- library(ggwordcloud) -->
<!-- library(tm) -->
<!-- library(dplyr) -->
<!-- library(RColorBrewer) -->

<!-- #Create a vector containing only the text -->
<!-- gtext <- HMRF$Project.Title -->
<!-- # Create a corpus -->
<!-- gdocs <- Corpus(VectorSource(gtext)) -->
<!-- gdocs <- gdocs %>% -->
<!--   tm_map(removePunctuation) %>% -->
<!--   tm_map(stripWhitespace) -->
<!-- gdocs <- tm_map(gdocs, content_transformer(tolower)) -->
<!-- gdocs <- tm_map(gdocs, removeWords, stopwords("english")) -->

<!-- gdtm <- TermDocumentMatrix(gdocs) -->
<!-- gmatrix <- as.matrix(gdtm) -->
<!-- gwords <- sort(rowSums(gmatrix),decreasing=TRUE) -->
<!-- gdf <- data.frame(word = names(gwords),freq=gwords) -->

<!-- set.seed(1235) # for reproducibility -->
<!-- gp<-ggwordcloud(words = gdf$word, freq = gdf$freq, min.freq = 1,max.words=200, random.order=FALSE, rot.per=0.35,            colors=brewer.pal(8, "Dark2")) -->
<!-- #gp -->

<!-- #Create a vector containing only the text -->
<!-- pmtext <- pubmed$Title -->
<!-- # Create a corpus -->
<!-- pmdocs <- Corpus(VectorSource(pmtext)) -->
<!-- pmdocs <- pmdocs %>% -->
<!--   tm_map(removePunctuation) %>% -->
<!--   tm_map(stripWhitespace) -->
<!-- pmdocs <- tm_map(pmdocs, content_transformer(tolower)) -->
<!-- pmdocs <- tm_map(pmdocs, removeWords, stopwords("english")) -->

<!-- pmdtm <- TermDocumentMatrix(pmdocs) -->
<!-- pmmatrix <- as.matrix(pmdtm) -->
<!-- pmwords <- sort(rowSums(pmmatrix),decreasing=TRUE) -->
<!-- pmdf <- data.frame(word = names(pmwords),freq=pmwords) -->

<!-- set.seed(1235) # for reproducibility -->
<!-- pm<-ggwordcloud(words = pmdf$word, freq = pmdf$freq, min.freq = 1,max.words=200, random.order=FALSE, rot.per=0.35,            colors=brewer.pal(8, "Dark2")) -->
<!-- grid.arrange(arrangeGrob(pm, top = 'Pubmed'),arrangeGrob(gp, top = 'Grants'),nrow=2,heights=c(6,6)) -->


<!-- ``` -->

#### Compare grant funding with incidence and mortality

```{r grants versus incidence and mortality,  fig.height=8, fig.width=10}

if (!require("ggrepel")) install.packages("ggrepel")
if (!require("ggplot2")) install.packages("gglot2")
if (!require("gridExtra")) install.packages("gridExtra")
library(ggrepel)
library(ggplot2)
library(gridExtra)


grantpmsum<- read.delim("https://github.com/StatBiomed/BMDS-book/raw/main/notebooks/module5-epidemi/Grants_pubmed_summary.txt", sep = "\t", header = TRUE, stringsAsFactor=FALSE, check.names = TRUE)
grantpmsum

HKCancer_inc <- HKCancer[HKCancer$Type=='incidence',]
HKCancer_inc_sum <- data.frame(incidence=colSums(HKCancer_inc[,-(1:4)]))

HKCancer_mort <- HKCancer[HKCancer$Type=='mortality',]
HKCancer_mort_sum <- data.frame(mortality=colSums(HKCancer_inc[,-(1:4)]))

HKCancer_compare <-data.frame(cancer=grantpmsum$Cancer, grants=grantpmsum$Grants, pubmed=grantpmsum$Pubmed, incidence=colSums(HKCancer_inc[,-(1:4)]), mortality=colSums(HKCancer_mort[,-(1:4)]))

gp<-ggplot(HKCancer_compare, aes(y=grants, x=pubmed)) + geom_point() + scale_color_brewer(palette="Dark2") + theme_classic() +  geom_label_repel(data=HKCancer_compare, aes(label=cancer),  nudge_x = 2, size = 3.5,label.size = NA ,min.segment.length =unit(0, 'lines'), max.overlaps = 45, box.padding = 0.4) + stat_cor(method = "pearson", label.x = 3, label.y = 90)

pinc<-ggplot(HKCancer_compare, aes(y=pubmed, x=incidence)) + geom_point() + scale_color_brewer(palette="Dark2") + theme_classic() +  geom_label_repel(data=HKCancer_compare, aes(label=cancer),  nudge_x = 2, size = 3.5,label.size = NA ,min.segment.length =unit(0, 'lines'), max.overlaps = 15, box.padding = 0.5)+ stat_cor(method = "pearson", label.x = 40000, label.y = 0)

pmor<-ggplot(HKCancer_compare, aes(y=pubmed, x=mortality)) + geom_point() + scale_color_brewer(palette="Dark2") + theme_classic() +  geom_label_repel(data=HKCancer_compare, aes(label=cancer),  nudge_x = 2, size = 3.5,label.size = NA ,min.segment.length =unit(0, 'lines'), max.overlaps = 15, box.padding = 0.5) + stat_cor(method = "pearson", label.x = 40000, label.y = 0)


ginc<-ggplot(HKCancer_compare, aes(y=grants, x=incidence)) + geom_point() + scale_color_brewer(palette="Dark2") + theme_classic() +  geom_label_repel(data=HKCancer_compare, aes(label=cancer),  nudge_x = 2, size = 3.5,label.size = NA ,min.segment.length =unit(0, 'lines'), max.overlaps = 40, box.padding = 0.4) + stat_cor(method = "pearson", label.x = 40000, label.y = 0)

gmor<-ggplot(HKCancer_compare, aes(y=grants, x=mortality)) + geom_point() + scale_color_brewer(palette="Dark2") + theme_classic() +  geom_label_repel(data=HKCancer_compare, aes(label=cancer),  nudge_x = 2, size = 3.5,label.size = NA ,min.segment.length =unit(0, 'lines'), max.overlaps = 40, box.padding = 0.4) + stat_cor(method = "pearson", label.x = 40000, label.y = 0)

grid.arrange(gp,ginc,gmor,pinc, pmor, ncol=3)

```

### Open discussion

As a group discuss what cancer type would be most worthy of funding in Hong Kong. Discuss what statistics and figures could be used to support your decision. If possible also discuss other data/analyses that can be performed and/or other diseases that are also in need of funding in Hong Kong.
